---
title: "Vitamin C effect on Guinea Pig Tooth Growth"
author: "Jacques wagstaff"
date: "18 November 2016"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

For the second part of the project, we analyze the ToothGrowth data 
in the R datasets package. The data looks at the effect of vitamin C on tooth
growth in 60 guinea pigs. The data frame has 60 observations of 3 variables; 
the length of odontoblasts (cells responsible for tooth growth) (len), 
the Supplement type (supp) taking either ascorbic acid (VC) or orange juice (OJ), 
and the dose in milligrams/day taking one of (0.5, 1, and 2 mg/day). A basic 
summary of the data is given below:

## Loading the data and Basic Summary 
First we load and plot the data by groups, and obtain a summary to get an overview 
```{r, echo = TRUE, fig.width=5, fig.height=3, fig.align='center'}
data(ToothGrowth); dat<-ToothGrowth ;library(ggplot2)
dat$dose<-as.factor(dat$dose) # turn dose into factor variable
g <- ggplot(dat, aes(x = dose, y = len, group = supp))
g <- g + geom_point(size =10, pch = 21, colour="darkgreen" ,alpha = .5, aes(fill = supp))
g 
summary(ToothGrowth)
```

The aim of the project is to identify if the delivery Method and/or Dosage affect 
tooth grow in Guinea Pigs. We can already see some apparent effect of method
and dose from the plot above. We now quantify these effects. 

# Hypothesis testing

In this section we perform Hypothesis testing to compare tooth growth by supp 
and dose. There are 6 groups, labeled by one of method (OJ, VC) and one of 
dose (0.5,1,2). The Guinea Pigs are different in all groups, therefore we perform 
independent group analysis. We perform 2-group intervals for all possible group 
combinations, where the data is not paired (since the guinea pigs are different).

## Comparing method VS vs OJ
First we split the data into groups by dose, so that we can compare method VS vs OJ.
```{r}
dat05<-subset(dat, dose %in% c(0.5)) 
dat1<-subset(dat, dose %in% c(1))
dat2<-subset(dat, dose %in% c(2))
```
The Null hypothesis is that the difference in the means between the two
groups is zero. We perform an independent group t-test to test this hypothesis. The 
results for the p-values and 95% confidence intervals are presented in the table below:
```{r,echo=FALSE}
pvalsumm<-data.frame(dose=c(0.5,1,2),supp=c("OJ-VC","OJ-VC","OJ-VC"),
                     pValues=signif(c(
                             t.test(len~supp, paired=F, var.equal=F, data=dat05)$p.value, 
                             t.test(len~supp, paired=F, var.equal=F, data=dat1)$p.value, 
                             t.test(len~supp, paired=F, var.equal=F, data=dat2)$p.value), 3),
                     Interval1=signif(c(
                             t.test(len~supp, paired=F, var.equal=F, data=dat05)$conf.int[1], 
                             t.test(len~supp, paired=F, var.equal=F, data=dat1)$conf.int[1], 
                             t.test(len~supp, paired=F, var.equal=F, data=dat2)$conf.int[1]),3),
                     Interval2=signif(c(
                             t.test(len~supp, paired=F, var.equal=F, data=dat05)$conf[2], 
                             t.test(len~supp, paired=F, var.equal=F, data=dat1)$conf.int[2], 
                             t.test(len~supp, paired=F, var.equal=F, data=dat2)$conf.int[2]),3)
                             
                     )              
pvalsumm
```
The R-code for this table is given in the appendix.

We claim any p-value below $\alpha=0.05$ as significant. We observe small p-values 
for small dose levels (0.5 and 1). This supports the alternative hypothesis: 
true difference in means is not equal to 0. Hence, for these 
dose values the type of delivery method has a significant effect on the tooth 
growth length. More specifically, the method by orange juice "OJ" is significantly
more effective than the method by "VC" (ascorbic acid). Whereas for large dose (2), 
where the p-value is large and the interval contains 0, the delivery method has 
no effect i.e. both delivery methods are equally effective.

## Comparing dose groups
Next we subset by OJ method, and compare dose groups
```{r}
datOJ05_1<-subset(dat, supp %in% c("OJ") & dose %in% c("0.5","1"))
datOJ05_2<-subset(dat, supp %in% c("OJ") & dose %in% c("0.5","2"))
datOJ1_2<-subset(dat, supp %in% c("OJ") & dose %in% c("1","2"))
```
Finally we subset by VC method, and compare dose groups
```{r}
datVC05_1<-subset(dat, supp %in% c("VC") & dose %in% c("0.5","1"))
datVC05_2<-subset(dat, supp %in% c("VC") & dose %in% c("0.5","2"))
datVC1_2<-subset(dat, supp %in% c("VC") & dose %in% c("1","2"))
```
Again, the Null hypothesis is that the difference in the means between pairs
of groups is zero. The results for the p-values are presented in the table below:
The R-code for this table is given in the appendix.
```{r,echo=FALSE}
pvalsumm2<-data.frame(supp=c("OJ","OJ","OJ","VC","VC","VC"),
                      dose=c("0.5-1","0.5-2","1-2","0.5-1","0.5-2","1-2"),
                     pValues=signif(c(
                             t.test(len~dose, paired=F, var.equal=F, data=datOJ05_1)$p.value,
                             t.test(len~dose, paired=F, var.equal=F, data=datOJ05_2)$p.value,
                             t.test(len~dose, paired=F, var.equal=F, data=datOJ1_2)$p.value,
                             t.test(len~dose, paired=F, var.equal=F, data=datVC05_1)$p.value,
                             t.test(len~dose, paired=F, var.equal=F, data=datVC05_2)$p.value,
                             t.test(len~dose, paired=F, var.equal=F, data=datVC1_2)$p.value),3),
                     Interval1=signif(c(
                             t.test(len~dose, paired=F, var.equal=F, data=datOJ05_1)$conf.int[1],
                             t.test(len~dose, paired=F, var.equal=F, data=datOJ05_2)$conf.int[1],
                             t.test(len~dose, paired=F, var.equal=F, data=datOJ1_2)$conf.int[1],
                             t.test(len~dose, paired=F, var.equal=F, data=datVC05_1)$conf.int[1],
                             t.test(len~dose, paired=F, var.equal=F, data=datVC05_2)$conf.int[1],
                             t.test(len~dose, paired=F, var.equal=F, data=datVC1_2)$conf.int[1]),3),
                     Interval2=signif(c(
                             t.test(len~dose, paired=F, var.equal=F, data=datOJ05_1)$conf.int[2],
                             t.test(len~dose, paired=F, var.equal=F, data=datOJ05_2)$conf.int[2],
                             t.test(len~dose, paired=F, var.equal=F, data=datOJ1_2)$conf.int[2],
                             t.test(len~dose, paired=F, var.equal=F, data=datVC05_1)$conf.int[2],
                             t.test(len~dose, paired=F, var.equal=F, data=datVC05_2)$conf.int[2],
                             t.test(len~dose, paired=F, var.equal=F, data=datVC1_2)$conf.int[2]),3))
pvalsumm2
```
We observe small p-values $p<0.05$ for all combinations, and all negative intervals.
Hence an increase in dose results in a significant increase in the length of odontoblasts, for both
delivery methods. The p-values are generally smaller for the "VC" method, this 
tells us that the difference in effectiveness between the "VC" and "OJ" methods 
seems to get smaller as the dose increases. This can also be seen in the data plot above.

## Assumptions

For the statistical testing we carried out we assumed the following:

1. The underlying data is Normally distributed

2. We assumed unequal variances in the paired t-tests

3. We tested many hypotheses, and assumed that no correcting for this is required.
This is a fair assumption because most of our p-values are very small anyway. 

# Conclusions

Here we answer the question: Do delivery Method and/or Dosage affect tooth grow 
in Guinea Pigs?

1. An increase in dose results in a significant increase in the length of odontoblasts, 
for both delivery methods.

2. For dose values of 0.5 and 1 mg/day the type of delivery method has a significant 
effect on the tooth growth length. More specifically, the method by orange juice "OJ" 
is significantly more effective than the method by "VC" (ascorbic acid). Whereas 
for large dose (2 mg/day), the delivery method has no effect i.e. both delivery 
methods are equally effective.

3. Hence the effectiveness of the "VC" method increases more rapidly as dose 
increase than the "OJ" method.

# Appendix

Here is the R-code for table 1:
```{r,echo=T}
pvalsumm<-data.frame(dose=c(0.5,1,2),supp=c("OJ-VC","OJ-VC","OJ-VC"),
                     pValues=signif(c(
                             t.test(len~supp, paired=F, var.equal=F, data=dat05)$p.value, 
                             t.test(len~supp, paired=F, var.equal=F, data=dat1)$p.value, 
                             t.test(len~supp, paired=F, var.equal=F, data=dat2)$p.value), 3),
                     Interval1=signif(c(
                             t.test(len~supp, paired=F, var.equal=F, data=dat05)$conf.int[1], 
                             t.test(len~supp, paired=F, var.equal=F, data=dat1)$conf.int[1], 
                             t.test(len~supp, paired=F, var.equal=F, data=dat2)$conf.int[1]),3),
                     Interval2=signif(c(
                             t.test(len~supp, paired=F, var.equal=F, data=dat05)$conf[2], 
                             t.test(len~supp, paired=F, var.equal=F, data=dat1)$conf.int[2], 
                             t.test(len~supp, paired=F, var.equal=F, data=dat2)$conf.int[2]),3)
                             
                     )              
pvalsumm
```

Here is the R-code for table 2:
```{r,echo=T}
pvalsumm2<-data.frame(supp=c("OJ","OJ","OJ","VC","VC","VC"),
                      dose=c("0.5-1","0.5-2","1-2","0.5-1","0.5-2","1-2"),
                     pValues=signif(c(
                             t.test(len~dose, paired=F, var.equal=F, data=datOJ05_1)$p.value,
                             t.test(len~dose, paired=F, var.equal=F, data=datOJ05_2)$p.value,
                             t.test(len~dose, paired=F, var.equal=F, data=datOJ1_2)$p.value,
                             t.test(len~dose, paired=F, var.equal=F, data=datVC05_1)$p.value,
                             t.test(len~dose, paired=F, var.equal=F, data=datVC05_2)$p.value,
                             t.test(len~dose, paired=F, var.equal=F, data=datVC1_2)$p.value),3),
                     Interval1=signif(c(
                             t.test(len~dose, paired=F, var.equal=F, data=datOJ05_1)$conf.int[1],
                             t.test(len~dose, paired=F, var.equal=F, data=datOJ05_2)$conf.int[1],
                             t.test(len~dose, paired=F, var.equal=F, data=datOJ1_2)$conf.int[1],
                             t.test(len~dose, paired=F, var.equal=F, data=datVC05_1)$conf.int[1],
                             t.test(len~dose, paired=F, var.equal=F, data=datVC05_2)$conf.int[1],
                             t.test(len~dose, paired=F, var.equal=F, data=datVC1_2)$conf.int[1]),3),
                     Interval2=signif(c(
                             t.test(len~dose, paired=F, var.equal=F, data=datOJ05_1)$conf.int[2],
                             t.test(len~dose, paired=F, var.equal=F, data=datOJ05_2)$conf.int[2],
                             t.test(len~dose, paired=F, var.equal=F, data=datOJ1_2)$conf.int[2],
                             t.test(len~dose, paired=F, var.equal=F, data=datVC05_1)$conf.int[2],
                             t.test(len~dose, paired=F, var.equal=F, data=datVC05_2)$conf.int[2],
                             t.test(len~dose, paired=F, var.equal=F, data=datVC1_2)$conf.int[2]),3))
pvalsumm2
```
