---
title: "Investigating the Exponential Distribution"
author: "Jacques wagstaff"
date: "17 November 2016"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Overview
The exponential distribution $f(x)=\lambda e^{-\lambda x}$, for $x\geq 0$, is 
characterised by one parameter: the rate parameter $\lambda$. The theoretical 
mean of the distribution is given by $\mu=1/\lambda$ and the standard deviation 
is also $\sigma=1/\lambda$.

The Central Limit Theorem (CLT) states that 
$$Z_n\equiv\frac{\bar X_n - \mu}{\sigma / \sqrt{n}}=
\lambda\sqrt{n}\left(\bar X_n -1/\lambda\right)$$
has a distribution of a standard normal $N\left(\mu=0,\sigma=1\right)$, 
where $\bar X_n$ is the sample average, if $n$ is large enough.

We investigate the distribution of averages of 40 exponentials, $n=40$, 
and run a thousand simulations, $nosim=1000$, to find the distribution of $Z_n$. 
We set $\lambda = 0.2$ for all of the simulations. The R code for the simulation is given below.

In the code below, we sample $n\times nosim$ numbers from the exponential distribution to simulate our data. We then take the average in groups of $n=40$. The product
$Z_n=\frac{\bar X_n - \mu}{\sigma / \sqrt{n}}$ is calculated for each group and
the results are then plotted. 


```{r, fig.height=3, fig.width=5, fig.align='center'}
nosim<-1000; n<-40; lambda<-0.2; library(ggplot2); set.seed(1286)
cfunc <- function(x, n) sqrt(n)*(mean(x)-1/lambda)/(1/lambda)
# next we sample from the exponential distribution to simulate our data
dat <- data.frame(Z = apply(matrix(rexp(n*nosim,lambda),nosim), 1, cfunc, n))
g <- ggplot(dat, aes(x = Z)) + geom_histogram(alpha = .20, binwidth=.5,
                       colour = "darkgreen", fill = "blue", aes(y = ..density..)) 
g <- g + stat_function(fun = dnorm, size = 1,colour="red")
g<-g +geom_vline(xintercept = 0, colour="red") # The theoretical mean 1/lambda
g<-g +geom_vline(xintercept = mean(dat$Z)) # The sample mean 
g<-g+coord_cartesian(xlim = c(-4,4))
g
```

The plot above shows a histogram for the density of 
$Z_n=\frac{\bar X_n - \mu}{\sigma / \sqrt{n}}$. Overlaying the histogram we have 
also plotted the density function of a standard normal (line in red). The plot 
also shows two vertical lines showing the theoretical mean (in red) and the sample 
mean (in black).



## 1. The sample mean and the theoretical mean

For the sample mean and its interval we can perform a t.test:
```{r}
t.test(dat$Z)
```
The sample mean, also shown in the plot above as a vertical line in black, is given by
$-0.051$ which is very close to the expected theoretical mean of a standard normal 
distribution i.e. $0$. The 95% confidence interval for the mean is 
$\left(-0.112, 0.009\right)$, which includes $0$. Note that the t-test assumes iid Gaussian variables.



## 2. The sample variance and the theoretical variance

The sample variance is given by
```{r}
var(dat$Z)
```
which is very close to the expected theoretical variance of a standard normal 
distribution i.e. $1$.

## 3. Is the distribution Normal?

The first plot shown above, which shows a histogram for the density of 
$Z_n$ and an overlay of the density function of a standard normal (line in red), clearly shows that the distribution is to a good approximation a standard normal. This is also confirmed by the previous calculations of the mean and variance of the sample which are to a good approximation those of a standard normal.

To illustrate this point further, we show in the plot below the difference between the distribution of a large collection of random exponentials (non-avg) and the distribution of a large collection of averages of 40 exponentials (avg). 

```{r, fig.height=3, fig.width=9, fig.align='center'}
dat1 <- data.frame(x = c(apply(matrix(rexp(n*nosim,lambda),nosim), 1, mean),
        rexp(nosim,lambda)),
        type = factor(rep(c("avg", "non-avg"), rep(nosim, 2))))
g1 <- ggplot(dat1, aes(x = x, fill = type)) + 
        geom_histogram(alpha = .20, binwidth=.5, colour = "darkgreen",
                       aes(y = ..density..)) 
g1 + facet_grid(. ~ type)
```

Clearly the underlying distribution of the simulated data (non-avg) is not Gaussian (we know it comes from an exponential distribution), however the distribution of averages (avg) is clearly much more Gaussian.

## Conclusion

We conclude that: 
$\textbf{The distribution of means of 40 exponentials behave as predicted}\\$ 
$\textbf{by the Central Limit Theorem}$. 

